Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-04-08 index.qmd
4/8 update

Below is an example plot demonstrating scaled Shoenfeld partial
residuals plotted against time to test the proportional hazards
assumption. While there is some curvature in the line, the residuals are
roughly horizontal indicating proportional hazard over time for this
covariate.

@@

Though CPH is typically used with health data, it does have applications
in other fields. This paper will study the remaining useful life (RUL)
of a NASA aircraft engine using data provided by NASA for the study on
propagation modeling [@saxena2008]. In particular, this paper will be
examining the data from the engine make two which has a training dataset
and a testing dataset.

@@

Below are tables showing the structure of the training and testing data
after the transformations describd above.

@@

test <- read.delim(testurl, header = TRUE, sep = '') %>%
  mutate(status = 0)

kable(head(test, n = 5), format = "markdown", caption = "Testing Data")

@@

kable(head(train, n = 5), format = "markdown", caption = 'Training Data')
```

@@

There are 260 engines in the training data. The histogram shows that the
survival times of the engines are normally distributed and the box plot
shows there six outliers in the dataset. The engines have a median
survival time of 199 cycles with a standard deviation of 46.8. The
maximum number of cycles an engine lasted was 378 while the minimum
number of cycles an engined lasted was 128.

```{r, echo = FALSE}
distributions <- train %>% 
  group_by(id) %>%
  summarise(time = max(time))

summarymetrics <- data.frame(c("Minimum", "Median", "Mean", "Standard Deviation", "Maximum"),
                             c(min(distributions$time), 
                               median(distributions$time),
                               mean(distributions$time),
                               sd(distributions$time),
                               max(distributions$time)))

@@

```{r, echo=FALSE}
Histogram <- ggplot(data = distributions,
       aes(x = time)) +
  geom_histogram(binwidth = 5,
                 fill = "grey")+
  labs(x = "Survival Time (Iterations)",
       y = "Frequency")+
  theme_bw()

box <- ggplot(data = distributions,
                  aes(y = time))+
  geom_boxplot(fill = "gray",
               color = "black")+
  ylab("Time (Iterations)")+
  theme_bw()

@@

To determine the intitial survival function, the Kaplan-Meier survival
curve was generated and is shown below. The graph verifies that the
median survival time is roughly 200 cycles and shows the maximum cycle.
After that cycle has been reached, the survival probability plummets to
0%.

```{r}
t <- as.matrix(train[, "time"])
s <- as.matrix(train[, "status"])

@@

univ_models <- lapply( univ_formulas, function(x){coxph(x, data = train)})

@@

Based on the above test, the covariates that will be considered for the
CPH models are os1, os3, sm1, sm2, sm5, sm6, sm7, sm11, sm12, sm13,
sm14, sm19, sm20, and sm21 which is fourteen out of twenty-four possible
covariates.

@@

```{r}
cox1 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox2 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5)

cox3 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)

cox4 <- coxph(data = train, id = id,
             Surv(time, status) ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox5 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )

cox6 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1  + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)

cox7 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1  + sm13 + sm14 + sm19 + sm20)

cox8 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1 + os3 + sm21)

cox9 <- coxph(data = train, id = id,
             Surv(time, status) ~ os1 + os3 + sm2 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)

cox10 <- coxph(data = train, id = id,
             Surv(time, status) ~  os3 + sm1 + sm2 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)

@@

```{r}
par(mfrow = c(3,3),
      mar = c(2,2,4,1))

@@

plot(train$os1, martingale, main = "Martingale Residuals: os1")
abline(h = 0, col = "red")

plot(train$os2, martingale, main = "Martingale Residuals: os2")
abline(h = 0, col = "red")

plot(train$sm1, martingale, main = "Martingale Residuals: sm1")
abline(h = 0, col = "red")

plot(train$sm2, martingale, main = "Martingale Residuals: sm2")
abline(h = 0, col = "red")

plot(train$sm5, martingale, main = "Martingale Residuals: sm5")
abline(h = 0, col = "red")

plot(train$sm6, martingale, main = "Martingale Residuals: sm6")
abline(h = 0, col = "red")

plot(train$sm7, martingale, main = "Martingale Residuals: sm7")
abline(h = 0, col = "red")

plot(train$sm11, martingale, main = "Martingale Residuals: sm11")
abline(h = 0, col = "red")

plot(train$sm12, martingale, main = "Martingale Residuals: sm12")
abline(h = 0, col = "red")

plot(train$sm13, martingale, main = "Martingale Residuals: sm13")
abline(h = 0, col = "red")

plot(train$sm14, martingale, main = "Martingale Residuals: sm14")
abline(h = 0, col = "red")

plot(train$sm20, martingale, main = "Martingale Residuals: sm20")
abline(h = 0, col = "red")

plot(train$sm21, martingale, main = "Martingale Residuals: sm21")
abline(h = 0, col = "red")
```

@@

```{r}
cox1.1 <- coxph(data = train,Surv(time, status) ~ sm2 + sm5 + sm7 + sm11 + sm12 + sm20 + sm21)

kable(tidy(cox1.1), format = "markdown")
g <- glance(cox1)
g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
kable(g, format = "markdown")

@@

```{r}
m_2 <- cox.zph(cox1.1) %>%
  pluck("table") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  as_tibble()

kable(m_2, format = "markdown")

# Function to put ph_plots on one visual
create_ph_plots <- function(coxmodel, row, col){
  # Function takes argument that is a cox model from coxph()

@@

plot(train$sm2, martingale, main = "Martingale Residuals: sm2")
abline(h = 0, col = "red")

plot(train$sm5, martingale, main = "Martingale Residuals: sm5")
abline(h = 0, col = "red")

plot(train$sm7, martingale, main = "Martingale Residuals: sm7")
abline(h = 0, col = "red")

plot(train$sm11, martingale, main = "Martingale Residuals: sm11")
abline(h = 0, col = "red")

plot(train$sm12, martingale, main = "Martingale Residuals: sm12")
abline(h = 0, col = "red")

plot(train$sm20, martingale, main = "Martingale Residuals: sm20")
abline(h = 0, col = "red")

plot(train$sm21, martingale, main = "Martingale Residuals: sm21")
abline(h = 0, col = "red")

@@

```{r}
coef <- data.frame(CoefficientValue = cox1.1$coefficients)
kable(coef)
```

@@

```{r}
test$HazardRate <- predict(cox1.1,
        newdata = test,
        type = "risk")

predictiondata <- test %>%
  group_by(id) %>%
  slice_max(time) %>%
  mutate(CumulativeHazard = cumsum(HazardRate)) %>%
  select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, status, HazardRate)

kable(head(predictiondata %>% select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, HazardRate)))

@@

```{r}
kable(head(predictiondata))

@@

```{r}
predictiondata$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = predictiondata)
kable(head(predictiondata))

@@

#### Predicting Time Until Failure

Time until event can be modeled in R by passing the `type = "lp"`
argument in the `stats::predict()` function to get a value for linear
predictor.

@@

@@

-   Summarize your key findings.

-   Discuss the implications of your results.

>>>>

Below is an example plot demonstrating scaled Schoenfeld partial
residuals plotted against time to test the proportional hazards
assumption. While there is some curvature in the line, the residuals are
roughly horizontal indicating proportional hazard over time for this
covariate.

@@

Though CPH is typically used with health data, it does have applications
in other fields. This paper will study the survivability of a NASA
aircraft engine using data provided by NASA for the study on propagation
modeling [@saxena2008]. In particular, this paper will be examining the
data from the engine make two which has a training dataset and a testing
dataset.

@@

Below are tables showing the structure of the training and testing data
after the transformations described above.

@@

test <- read.delim(testurl, header = TRUE, sep = '') %>%
  mutate(status = 0,
         id = id + 260)

@@

data_all <- rbind(train, test)

kable(head(data_all, n = 5), format = "markdown", caption = 'NASA Aircraft Engine Data')
```

@@

There are 260 engines in the training data. The histogram shows that the
survival times of the engines are normally distributed and the box plot
shows there six outliers in the dataset. The engines have a median
survival time of 199 cycles with a standard deviation of 46.8. The
maximum number of cycles an engine lasted was 378 while the minimum
number of cycles an engine lasted was 128.

```{r, echo = FALSE}
train_grouped <- train %>% 
  group_by(id) %>%
  slice_max(time)

data_grouped <- data_all %>%
  group_by(id) %>%
  slice_max(time)

summarymetrics <- data.frame(c("Minimum", "Median", "Mean", "Standard Deviation", "Maximum"),
                             c(min(train_grouped$time), 
                               median(train_grouped$time),
                               mean(train_grouped$time),
                               sd(train_grouped$time),
                               max(train_grouped$time)))

@@

```{r, echo=FALSE}
Histogram <- ggplot(data = train_grouped,
       aes(x = time)) +
  geom_histogram(binwidth = 5,
                 fill = "grey")+
  labs(x = "Survival Time (Iterations)",
       y = "Frequency")+
  theme_bw()

box <- ggplot(data = train_grouped,
                  aes(y = time))+
  geom_boxplot(fill = "gray",
               color = "black")+
  ylab("Time (Iterations)")+
  theme_bw()

@@

To determine the initial survival function, the Kaplan-Meier survival
curve was generated and is shown below. The graph visualizes the
survival curve and the information that was gathered through the
histogram.

```{r}
t <- as.matrix(train_grouped[, "time"])
s <- as.matrix(train_grouped[, "status"])

@@

univ_models <- lapply( univ_formulas, function(x){coxph(x, data = data_grouped)})

@@

Based on the above test, the covariates that will be considered for the
CPH models are os1, os3, sm1, sm5, sm6, sm7, sm11, sm12, sm13, sm14,
sm19, sm20, and sm21 which is fourteen out of twenty-four possible
covariates.

@@

```{r}
surv_object <- Surv(data_grouped$time, data_grouped$status)

cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox2 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5)

cox3 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)

cox4 <- coxph(data = data_grouped, surv_object ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox5 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )

cox6 <- coxph(data = data_grouped, surv_object ~ os1  + sm1 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)

cox7 <- coxph(data = data_grouped, surv_object ~ os1  + sm13 + sm14 + sm19 + sm20)

cox8 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm21)

cox9 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)

cox10 <- coxph(data = data_grouped, surv_object ~  os3 + sm1 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)

@@

```{r}
par(mfrow = c(4,4),
      mar = c(2,2,4,1))

@@

plot(data_grouped$os1, martingale, main = "Martingale Residuals: os1")
abline(h = 0, col = "red")

plot(data_grouped$os2, martingale, main = "Martingale Residuals: os2")
abline(h = 0, col = "red")

plot(data_grouped$sm1, martingale, main = "Martingale Residuals: sm1")
abline(h = 0, col = "red")

plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
abline(h = 0, col = "red")

plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
abline(h = 0, col = "red")

plot(data_grouped$sm6, martingale, main = "Martingale Residuals: sm6")
abline(h = 0, col = "red")

plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
abline(h = 0, col = "red")

plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
abline(h = 0, col = "red")

plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
abline(h = 0, col = "red")

plot(data_grouped$sm13, martingale, main = "Martingale Residuals: sm13")
abline(h = 0, col = "red")

plot(data_grouped$sm14, martingale, main = "Martingale Residuals: sm14")
abline(h = 0, col = "red")

plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
abline(h = 0, col = "red")

plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
abline(h = 0, col = "red")
```

@@

```{r}
cox1.1 <- coxph(data = data_grouped,Surv(time, status) ~ sm2 + sm5 + sm7 + sm11 + sm12 + sm20 + sm21)

kable(tidy(cox1.1), format = "markdown")
g <- glance(cox1.1)
g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
kable(g, format = "markdown")

@@

```{r}
ph_2 <- cox.zph(cox1.1) %>%
  pluck("table") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  as_tibble()

kable(ph_2, format = "markdown")

@@

plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
abline(h = 0, col = "red")

plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
abline(h = 0, col = "red")

plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
abline(h = 0, col = "red")

plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
abline(h = 0, col = "red")

plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
abline(h = 0, col = "red")

plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
abline(h = 0, col = "red")

plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
abline(h = 0, col = "red")

@@

```{r}
coef <- data.frame(CoefficientValue = cox1.1$coefficients)
kable(coef, format = "markdown")
```

@@

```{r}
data_grouped_reduced <- data_grouped %>% select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, status)

data_grouped_reduced$HazardRate <- predict(cox1.1,
        newdata = data_grouped_reduced,
        type = "risk")

@@

The cumulative hazard of individuals in the model are shown in the visual below. Higher cumulative hazard indicates a higher level of risk at time `t`.

```{r}
sf <- survfit(cox1.1)

ggsurvplot(fit = sf,
           data = data_grouped_reduced, risk.table = TRUE, fun = "cumhaz")$plot +
  ggtitle('Cumulative Hazard Over Time')

@@

```{r}
data_grouped_reduced$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = data_grouped_reduced)
kable(head(data_grouped_reduced))

@@

The resulting survival curve from the model is shown below.

@@

survtimes <- data.frame(Percent = c("100% Survival", "< 25% Survival,", "< 50% Survival", "< 75% Survival", "<90% Survival"),
                        Time = c(sf$time[sf$lower < 1][1], 
                                    sf$time[sf$lower < .75][1], 
                                    sf$time[sf$lower < .5][1], 
                                    sf$time[sf$lower < .25][1], 
                                    sf$time[sf$lower < .1][1]))

ggsurvplot(fit = sf, data = data_grouped_reduced)$plot +
  ggtitle('Survival Probability Over Time')

@@

The implications of an engine failure are severe, with a possible result including an aircraft crash with a loss of human life and millions of dollars. Based on these implications, there is no acceptable room for failure for aircraft engines. The model created a confidence interval for estimating probabilities at each value of time. The values for time used will be the value where the probability of survival at the lower end of the confidence interval to exercise caution.

```{r}
kable(survtimes)

The lower end of the confidence interval for survival probability is shown to hold stable at 100% until `t = 128` with a gradual decrease afterwards. Half of the engines are expected to survive until `t = 221` and 90% of the engines are expected to fail at `t = 316`. These times are reflected by the survival curve shown above. The cumulative hazard over time plot confirms the hazard over time increases. The plot shows a low hazard until `t = 128` with a gradual increase.

Based on these findings, the model can be applied by recommending for NASA not to use engines after 128 iterations have passed. The engines should undergo routine maintenance beforehand and should be evaluated to ensure the probability of survival calculated by the model remains at 100%. All engines should be equipped with sensors for the covariates used and readings should be applied to the model after each use of the engine. The engine should be removed from service if the resulting survival probability is less than 100%.
